---
title: "Final Project"
author: "Nakoul Makim (namakim)"
date: "3/22/2020"
output: html_document
---

## R Program's external requirements
```{r}
library(tidyverse)
library(caret)
library(dplyr)
library(ggplot2)
library(standardize)
library(nnet)
library(glmnet)
library(arm)
library(multcomp)
```

## Data Cleaning
```{r}
#at-bat data
at_bats <- read_csv("atbats.csv")
#at_bats

#pitch data
pitches <- read_csv("pitches.csv")
#pitches

#player name data
player_names <- read_csv("player_names.csv")

#combine first and last name of players
player_names$player_name <- paste(player_names$first_name, player_names$last_name)
players <- subset(player_names, select = c("id","player_name"))
#players
```

```{r}
#trimmed datasets
at_bats_trimmed <- subset(at_bats, select = c("ab_id", "batter_id", "pitcher_id", "inning", "o", "p_score", "p_throws", "stand"))
#at_bats_trimmed

pitches_trimmed <- subset(pitches, select = c("ab_id","end_speed", "px", "pz", "pitch_type","b_count","s_count","outs","pitch_num","on_1b","on_2b","on_3b"))
#pitches_trimmed
```

```{r}
# match the column names of the 2 tables in order to join
names(at_bats_trimmed) [names(at_bats_trimmed) == "batter_id"] <- "id"
names(at_bats_trimmed) [names(at_bats_trimmed) == "pitcher_id"] <- "id_p"

# joining batter and player name data
combined <- full_join(at_bats_trimmed, players, "id")

# joining pitcher and player name data
names(players)[names(players) == "id"] <- "id_p"
combined_player_data <- full_join(combined, players, "id_p")
#combined_player_data
```

```{r}
# Trimming the data more, removed batter name, player ID's, score data
clean_data <- subset(combined_player_data, select = c("ab_id","player_name.y","inning","o","p_throws","stand"))

# combined pitch data with batter data
combined_bat_pitch <- full_join(clean_data, pitches_trimmed, "ab_id")
#combined_bat_pitch
```

```{r}
# added year column
combined_bat_pitch$year <- strtrim(combined_bat_pitch$ab_id, 4)

# removed data from 2015 and 2016 years due to large amounts of missing data in those years
final_data <- subset(combined_bat_pitch, year %in% c("2017","2018"))
names(final_data)[names(final_data) == "player_name.y"] = "pitcher_name"
#final_data
```

```{r}
# read in additional pitching stats for each pitcher from 2017-2019
pitcher_stats <- read_csv("stats.csv")

# combine first and last name of pitchers
pitcher_stats$pitcher_name <- paste(pitcher_stats$first_name, pitcher_stats$last_name)

# make pitch-type column names more readable and easier to understand
names(pitcher_stats) [names(pitcher_stats) == "n_ff_formatted"] <- "4-seam %"
names(pitcher_stats) [names(pitcher_stats) == "n_sl_formatted"] <- "slider %"
names(pitcher_stats) [names(pitcher_stats) == "n_ch_formatted"] <- "change-up %"
names(pitcher_stats) [names(pitcher_stats) == "n_cukc_formatted"] <- "curve ball %"
names(pitcher_stats) [names(pitcher_stats) == "n_sift_formatted"] <- "sinker %"
names(pitcher_stats) [names(pitcher_stats) == "n_fc_formatted"] <- "cutter %"
names(pitcher_stats) [names(pitcher_stats) == "n_fs_formatted"] <- "splitter %"
names(pitcher_stats) [names(pitcher_stats) == "n_fastball_formatted"] <- "fastball %"
names(pitcher_stats) [names(pitcher_stats) == "n_breaking_formatted"] <- "breaking %"
names(pitcher_stats) [names(pitcher_stats) == "n_offspeed_formatted"] <- "offspeed %"

# when a pitcher cannot throw certain type of pitch, replace NA with 0
pitcher_stats[is.na(pitcher_stats)] = 0

# Trim the additional pitch stats 
pitcher_stats_trimmed <- subset(pitcher_stats, select = c("pitcher_name","year","p_ball","p_called_strike","exit_velocity_avg","4-seam %","slider %","change-up %","curve ball %","sinker %","cutter %","splitter %","fastball %","breaking %","offspeed %","pitch_count"))
#pitcher_stats_trimmed
```

```{r}
# seperate dataset by year: 2017, 2018, 2019
pitcher_2017 <- subset(pitcher_stats_trimmed, year %in% "2017")
pitcher_2018 <- subset(pitcher_stats_trimmed, year %in% "2018")
pitcher_2019 <- subset(pitcher_stats_trimmed, year %in% "2019")

pitch_2017 <- subset(final_data, year %in% "2017")
pitch_2018 <- subset(final_data, year %in% "2018")
pitch_2019 <- subset(final_data, year %in% "2019")


combined_2017 <- full_join(pitch_2017, pitcher_2017, "pitcher_name")
combined_2018 <- full_join(pitch_2018, pitcher_2018, "pitcher_name")
combined_2017$ab_id <- substr(combined_2017$ab_id, 5, 10)
combined_2018$ab_id <- substr(combined_2018$ab_id, 5, 10)
#combined_2017
#combined_2018
```

```{r}
baseball_temp <- rbind(combined_2017, combined_2018)
baseball <- subset(baseball_temp, select =-c(year.y))
baseball$event <- NULL
baseball$pitch_count <- NULL
baseball$year.y <- NULL
baseball$p_ball <- NULL
baseball$p_called_strike <- NULL
baseball$on_base <- baseball$on_1b + baseball$on_2b + baseball$on_3b
baseball$on_1b <- NULL
baseball$on_2b <- NULL
baseball$on_3b <- NULL
baseball$p_throws <- NULL
baseball$stand <- NULL
baseball$spin_rate <- NULL
baseball$spin_dir <- NULL
baseball$type_confidence <- NULL
baseball$nasty <- NULL
baseball$code <- NULL
baseball$type <- NULL
baseball$exit_velocity_avg <- NULL
options(digits = 10)
names(baseball)[names(baseball) == "year.x"] = "Year"
baseball$pitch_type <- factor(baseball$pitch_type)
#names(baseball)[names(baseball) == "end_speed"] = "prev_pitch_speed"
#names(baseball)[names(baseball) == "px"] = "prev_x_loc"
#names(baseball)[names(baseball) == "pz"] = "prev_z_loc"
baseball$px <- round(baseball$px, 3)
baseball$pz <- round(baseball$pz, 3)
#baseball
baseball_2018 <- subset(baseball, baseball$Year == "2018")
#baseball_2018[complete.cases(baseball_2018),]
```

```{r}
baseball <- mutate(baseball, prev_pitch_speed = lag(end_speed))
baseball <- mutate(baseball, prev_x_loc = lag(px))
baseball <- mutate(baseball, prev_z_loc = lag(pz))
baseball_no_na <- na.omit(baseball)
#baseball
baseball_no_na <- baseball_no_na[!(baseball_no_na$pitch_type %in% c("EP","FO","IN","PO","SC","UN","AB")),]
baseball_no_na %>% mutate_if(is.factor, as.character) -> baseball_no_na
baseball_no_na$pitch_type[baseball_no_na$pitch_type %in% c("FF","FT")] <- "Fastball"
baseball_no_na$pitch_type[baseball_no_na$pitch_type %in% c("FC","SI")] <- "Modified Fastball"
baseball_no_na$pitch_type[baseball_no_na$pitch_type %in% c("SL","CU","KC","KN")] <- "Curveball"
baseball_no_na$pitch_type[baseball_no_na$pitch_type %in% c("CH","FS")] <- "Offspeed"
baseball_no_na$pitch_type <- factor(baseball_no_na$pitch_type)
#baseball_no_na
```

```{r}
mlb_temp <- subset(baseball_no_na, baseball_no_na$inning %in% c(1, 2, 3, 4, 5, 6, 7, 8, 9))
mlb_temp$count <- paste0(mlb_temp$b_count, mlb_temp$s_count)
mlb_temp$count <- as.numeric(mlb_temp$count)
mlb_temp$count <- as.factor(mlb_temp$count)
mlb_temp$inning <- as.factor(mlb_temp$inning)
mlb_temp$o <- as.factor(mlb_temp$o)
#mlb_temp
```
```{r}
mlb <- subset(mlb_temp, pitcher_name %in% c("Max Scherzer","Aaron Nola","Jacob deGrom","Corey Kluber","Patrick Corbin","Jake Arrieta","Justin Verlander","Trevor Bauer","Zack Greinke","Carlos Carrasco"))
#mlb
mlb_2017 <- subset(mlb, Year == "2017")
mlb_2017$pitcher_name <- as.factor(mlb_2017$pitcher_name)
#mlb_2017
mlb_2018 <- subset(mlb, Year == "2018")
mlb_2018$pitcher_name <- as.factor(mlb_2018$pitcher_name)
#mlb_2018
```

## Data Exploration
```{r}
ggplot(mlb, aes(x = pitch_type)) + geom_bar(fill = "steelblue") + geom_text(aes(label = stat(count)), stat = 'count', vjust = 1.6, color = "white", size = 3.5) + theme_minimal() + labs(y = "Frequency", x = "Pitch Type", title = "Number of Types of Pitches in Dataset")
```


```{r}
fastball <- subset(mlb, mlb$pitch_type == "Fastball")
curveball <- subset(mlb, mlb$pitch_type == "Curveball")
modified_fast <- subset(mlb, mlb$pitch_type == "Modified Fastball")
offspeed <- subset(mlb, mlb$pitch_type == "Offspeed")
```

## Methodology & Models:

```{r}
mlb_glm <- rbind(fastball, curveball)
mlb_glm$pitch_type <- droplevels(mlb_glm$pitch_type)
train_temp <- subset(mlb_glm, Year == "2017")
test_temp <- subset(mlb_glm, Year == "2018")
x <- which(colnames(train_temp) %in% c("4-seam %", "slider %", "change-up %", "curve ball %", "sinker %", "cutter %", "splitter %", "fastball %", "breaking %", "offspeed %"))
train <- train_temp[,-x]
y <- which(colnames(test_temp) %in% c("4-seam %", "slider %", "change-up %", "curve ball %", "sinker %", "cutter %", "splitter %", "fastball %", "breaking %", "offspeed %"))
test <- test_temp[,-y]
ggplot(mlb_glm, aes(x = pitch_type)) + geom_bar(fill = "steelblue") + geom_text(aes(label = stat(count)), stat = 'count', vjust = 1.6, color = "white", size = 3.5) + theme_minimal() + labs(x = "Pitch Type", y = "count", title = "Number of Fastballs and Curveball")
ggplot(mlb_glm, aes(x = pitch_type, y = prev_pitch_speed)) + geom_boxplot() + coord_flip() + stat_summary() + labs(x = "Previous Pitch Speed", y = "Pitch Type", title = "Boxplot of Previous Pitch Speeds of a Pitch Type")
ggplot(mlb_glm, aes(x = pitch_type, y = prev_x_loc)) + geom_boxplot() + labs(x = "Pitch Type", y = "Previous Horizaonal Location", title = "Boxplot of Previous pitch location of a Pitch Type")
ggplot(mlb_glm, aes(prev_pitch_speed, y = prev_x_loc, color = pitch_type)) + geom_point() + labs(x = "Previous pitch speed", y = "Previous x location", title = "Scatterplot of pitch speed vs pitch location")
ggplot(mlb_glm, aes(count, fill = pitch_type)) + geom_bar() + labs(x = "Batter Count", y = "Frequency", title = "The Type of Pitch thrown for each Batter Count")
```

### More Data Exploration of trimmed dataset
```{r}
with(mlb_glm, table(count, pitch_type))
```

### Binary Logistic Regression Model
```{r}
#Binary Logistic Regression
glm_test2 <- glm(pitch_type ~ log(prev_pitch_speed) + prev_x_loc + prev_z_loc + count + o + inning, family = "binomial", data = mlb_glm)
summary(glm_test2)
```

### Lasso Regression (Dimension Reduction)
```{r}
X4 <- model.matrix(glm_test2)[,-1]
y4 <- mlb_glm$pitch_type
lasso_mod4 <- glmnet(x = X4, y = y4, standardize = TRUE, family = "binomial")
plot(lasso_mod4, xvar = "dev", label = TRUE)
```

```{r}
lasso_cv4 <- cv.glmnet(x = X4, y = y4, family = "binomial", type.measure = "class")
plot(lasso_cv4)
best_lambda4 <- lasso_cv4$lambda.min
best_lambda4
lasso_best4 <- glmnet(x = X4, y = y4, family = "binomial", lambda = best_lambda4)
```

```{r}
pred4 <- predict(lasso_best4, s = best_lambda4, newx = X4, type = "class")
final_pred4 <- cbind(levels(y4)[y4], pred4)
head(final_pred4, 30)
coef(lasso_best4)
```

```{r}
df4 <- data.frame(levels(y4)[y4], pred4)
names(df4) <- c("Actual", "Predicted")
#df4
df_truth4 <- subset(df4, df4$Actual == df4$Predicted)
#df_truth4

Lasso_prediction_error <- 1 - (nrow(df_truth4)/ nrow(df4))
Lasso_prediction_error
```

### Fitting the Reduced-GLM and Prediction
```{r}
train$pitcher_name <- as.factor(train$pitcher_name)
test$pitcher_name <- as.factor(test$pitcher_name)

# New GLM Model with Updated Predictors
glm_model <- glm(pitch_type ~ prev_x_loc + log10(prev_pitch_speed) + count + factor(o, exclude = c("1")) + inning, family = "binomial", data = train)
#summary(glm_model)
pred <- predict(glm_model, test, type = "response")
pred <- na.omit(round(pred))
pred1 <- mean(pred)

# subtract 1 because in test data, curveball level = 1 and fastball = 2 so it is on 1-2 scale, not 0-1 scale like our prediction model. 
actual <- mean(as.numeric(test$pitch_type)) - 1
bias_pred <- pred1 - actual
pred_df <- data.frame("Mean", pred1, actual)
names(pred_df) <- c("","Prediction", "Actual")
bias_df <- data.frame("Bias", abs(bias_pred), "NA")
names(bias_df) <- c("", "Prediction", "Actual")
pred_act_df <- rbind(pred_df, bias_df)
knitr::kable(pred_act_df)
```

### Monte Carlo Simulation #1
```{r}
set.seed(3343229)
runs <- 23454
mc_func <- function() {
    # 0 = curveball, 1 = fastball
    z <- sample(c(0,1),46786, prob = c(0.44, 0.56), replace = TRUE)
    return(mean(z))
}
mc.sim <- replicate(runs, mc_func())
```

### Mean, Bias, Variance, and MSE of Simulation 1
```{r}
mean(mc.sim)
var(mc.sim)
test_df <- as.data.frame(mc.sim)
test_df2 <- mutate(
    test_df,
    test_df$pitch_type <- ifelse(test_df$mc.sim < 0.56, "Curveball", "Fastball")
)
names(test_df2) <- c("MC.Sim Value", "Pitch Type")
#test_df2
theta <- mean(as.numeric(test$pitch_type))
theta_hat <- mean(as.numeric(as.factor(test_df2$`Pitch Type`)))
bias <- theta_hat - theta
bias
MSE <- var(mc.sim) + bias^2
MSE
```


### Monte Carlo Simulation #2
```{r}
set.seed(3343229)
mean_prev_xloc <- mean(mlb_glm$prev_x_loc)
sd_xloc <- sqrt(var(mlb_glm$prev_x_loc))

mean_pitch_speed <- mean(mlb_glm$prev_pitch_speed)
sd_pitch_speed <- sqrt(var(mlb_glm$prev_pitch_speed))

runs <- 23454
mc_func2 <- function(){
    # count variables where 0 = "0",....,3 = "10",....12 = "32" 
    count <- sample(c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12), 23454, replace = TRUE)
    # innings
    innings <- sample(c(1, 2, 3, 4, 5, 6, 7, 8, 9), 23454, replace = TRUE)
    
    outs <- sample(c(0, 1, 2, 3), 23454, replace = TRUE)
    
    x_loc <- rnorm(23454, mean = mean_prev_xloc, sd = sd_xloc)
    
    prev_speed <- rnorm(23454, mean = mean_pitch_speed, sd = sd_pitch_speed)
    
    pitch_class <- 0.229996*count + -0.139265*outs + -0.446322*innings + -0.033818*x_loc + -0.042988339*prev_speed
    
    if (mean(pitch_class) <= -4.44335) {
        pitch_class <- 0
    }
    else {
        pitch_class <- 1
    }
    return(mean(pitch_class))
}

mc.sim2 <- replicate(runs, mc_func2())
```

### Mean, Variance, Bias, MSE of Simulation 2
```{r}
mean(mc.sim2)
var(mc.sim2)
test_df3 <- as.data.frame(mc.sim2)
#test_df3
test_df4 <- mutate(
    test_df3,
    test_df3$pitch_type <- ifelse(test_df3$mc.sim2 == 0, "Curveball", "Fastball")
)
names(test_df4) <- c("MC.Sim Value", "Pitch Type")
#test_df2
theta <- mean(as.numeric(test$pitch_type))
theta_hat2 <- mean(as.numeric(as.factor(test_df4$`Pitch Type`)))
bias2 <- theta_hat2 - theta
bias2
MSE2 <- var(mc.sim2) + bias2^2
MSE2
```

### Simulation Results
```{r}
MC_df <- data.frame("Simulation 1", mean(mc.sim), var(mc.sim), abs(bias), MSE)
names(MC_df) <- c("", "Mean of Simulation","Variance of Simulation", "Bias of Simulation", "Mean Squared Error")
MC_df2 <- data.frame("Simulation 2", mean(mc.sim2), var(mc.sim2), abs(bias2), MSE2)
names(MC_df2) <- c("","Mean of Simulation","Variance of Simulation", "Bias of Simulation", "Mean Squared Error")
MC <- rbind(MC_df, MC_df2)
knitr::kable(MC, main = "Table of Mean and Variance of Monte-Carlo Simulations")
```

